---
title: "Clustering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r ,echo=FALSE,message = FALSE,warning = FALSE}
############### 0. load packages --------------------------
library(data.table)
library(magrittr)
library(mapview)
library(ggplot2)
library(readr)
library(factoextra)
library(patchwork)
library(NbClust)


options(scipen = 999)
`%nin%` <- Negate(`%in%`)

```

# 1. Prep data

The cols I used was
```{r ,echo=TRUE,warning=FALSE}
df_raw <- readr::read_rds("../../../../data/urbanformbr/pca_regression_df/pca_regression_df_ready_to_use.rds")
# select columns
df_prep <- data.table::copy(df_raw)[,.SD
                                    ,.SDcols = c("i_code_urban_concentration"
                                                 , "i_name_urban_concentration"
                                                 , "i_name_uca_case"
                                                 , "x_proportion_largest_patch"
                                                 , "x_density_pop_05km_total_2014"
                                                 , "x_avg_cell_distance"
                                                 , "x_land_use_mix"
                                                 #,"x_pop_2015"
                                    )]
```

The method to normalize was
\begin{equation}
x_{norm} = log_{10}\big(x + \sqrt{x^2+1}\big)
\end{equation}

```{r ,echo=FALSE}
cols_to_norm1 <-  c("x_proportion_largest_patch"
                    , "x_density_pop_05km_total_2014"
                    , "x_avg_cell_distance"
                    , "x_land_use_mix"
                    , "x_pop_2015")
# normalize
cols_to_norm <- base::setdiff(names(df_prep),c("i_code_urban_concentration",
                                               "i_name_urban_concentration",
                                               "i_name_uca_case"))

my_f <- function(i){log10(i + sqrt(i^2 + 1) )}
df_prep[, (cols_to_norm) := lapply(.SD, as.numeric), .SDcols = cols_to_norm]
#df_prep[, (cols_to_norm) := lapply(.SD, log10), .SDcols = cols_to_norm]
df_prep[, (cols_to_norm) := lapply(.SD, my_f), .SDcols = cols_to_norm]
df_prep <- df_prep[,.SD,.SDcols = cols_to_norm]
```

# 2. clustering tendency

Initially, let's verify if there's clustering tendency. 
We can compare with random data


```{r ,echo=TRUE}
# Random data generated from the iris data set
random_df <- apply(df_prep, 2,
                   function(x){runif(length(x), min(x), (max(x)))}) %>%
  as.data.frame()
# Visual ilustration
p01 <- factoextra::fviz_pca_ind(prcomp(df_prep), title = "PCA - Urban form data",
                                #habillage = df_raw$i_code_urban_concentration,
                                palette = "jco",
                                geom = "point", ggtheme = theme_classic(),
                                legend = "bottom")
# Plot the random df
p02 <- factoextra::fviz_pca_ind(prcomp(random_df), title = "PCA - Random data",
                                geom = "point", ggtheme = theme_classic())

p01 | p02
```

some other visual methods includes and ordered distance matrix


```{r ,echo=TRUE}
p3 <- factoextra::fviz_dist(dist(df_prep), show_labels = FALSE)+
  labs(title = "Urban form data")
p4 <- factoextra::fviz_dist(dist(random_df), show_labels = FALSE)+
  labs(title = "Random data")

p3 | p4
```
The dissimilarity matrix image confirms that there is a cluster structure in the 
iris data set but not in the random one.


# 3. optimal number of clustering 

**Elbow_method**

Define clusters such that the total intra-cluster variation (or total
within-cluster sum of square (WSS)) is minimized. The total WSS measures the
compactness of the clustering and we want it to be as small as possible

```{r ,echo=TRUE}
p_31 <- factoextra::fviz_nbclust(df_prep, kmeans, method = "wss") +
  labs(subtitle = "Elbow method")

p_31
```

**Silhouette_method**

Briefly, it measures the quality of a clustering. That is, it determines how
well each object lies within its cluster. A high average silhouette width
indicates a good clustering. Average silhouette method computes the average
silhouette of observations for different values of k.
The optimal number of clusters k is the one that maximize the average
silhouette over a range of possible values for k (Kaufman and Rousseeuw [1990]).
```{r ,echo=TRUE}
p_32 <- factoextra::fviz_nbclust(df_prep, kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method")
```

**Gap_statistic_method**

The gap statistic compares the total within intra-cluster variation for
different values of k with their expected values under null reference
distribution of the data. The estimate of the optimal clusters will be
value that maximize the gap statistic (i.e, that yields the largest gap
statistic). This means that the clustering structure is far away
from the random uniform distribution of points.

```{r ,echo=TRUE}
set.seed(123)
c_33 <- NbClust::NbClust(df_prep, distance = "euclidean", min.nc = 2,
                         max.nc = 10, method = "complete",index = "gap")
p_33 <- factoextra::fviz_nbclust(x = df_prep, kmeans,
                                 nstart = 25, method = "gap_stat",
                                 nboot = 50)+
  labs(subtitle = "Gap statistic method")

p_31 + p_32 / p_33
```


*Some other methods*

```{r ,echo=TRUE}

n_31 <- NbClust::NbClust(data = df_prep,diss = NULL
                         , distance = "euclidean", min.nc = 2
                         , max.nc = 10,index = "all", method = "kmeans")

n_31
```


# 4. Viewing clustering 

### 4.1 - K-means clustering
```{r ,echo=TRUE}
km.res2 <- factoextra::eclust(x = df_prep
                              ,FUNcluster =  "kmeans"
                              , k = 2
                              , nstart = 25
                              , graph = FALSE)

factoextra::fviz_cluster(km.res2)
```

```{r ,echo=TRUE}
km.res3 <- factoextra::eclust(x = df_prep
                              ,FUNcluster =  "kmeans"
                              , k = 3
                              , nstart = 25
                              , graph = FALSE)
factoextra::fviz_cluster(km.res3)
```

```{r ,echo=TRUE}
km.res4 <- factoextra::eclust(x = df_prep
                              ,FUNcluster =  "kmeans"
                              , k = 4
                              , nstart = 25
                              , graph = TRUE)
```

### 4.2 - hierarchical 


```{r ,echo=TRUE}
hc.res2 <- factoextra::eclust(x = df_prep
                              , FUNcluster =  "hclust"
                              , k = 2
                              , hc_metric = "euclidean"
                              , hc_method = "ward.D2",
                              graph = FALSE)
factoextra::fviz_dend(hc.res2, show_labels = FALSE,
                      palette = "jco", as.ggplot = TRUE)
```
```{r ,echo=TRUE}
hc.res3 <- factoextra::eclust(x = df_prep
                              , FUNcluster =  "hclust"
                              , k = 3
                              , hc_metric = "euclidean"
                              , hc_method = "ward.D2",
                              graph = FALSE)
factoextra::fviz_dend(hc.res3, show_labels = FALSE,
                      palette = "jco", as.ggplot = TRUE)

hc.res4 <- factoextra::eclust(x = df_prep
                              , FUNcluster =  "hclust"
                              , k = 4
                              , hc_metric = "euclidean"
                              , hc_method = "ward.D2",
                              graph = FALSE)
factoextra::fviz_dend(hc.res4, show_labels = FALSE,
                      palette = "jco", as.ggplot = TRUE)
```

# 5 analysis

```{r ,echo=FALSE}

df_raw[,cluster2 := hc.res2$cluster]
df_raw[,cluster3 := hc.res3$cluster]
df_raw[,cluster4 := hc.res4$cluster]

df_plot <- data.table::melt(data = df_raw
                            ,id.vars = c('i_code_urban_concentration',
                                         cols_to_norm1,
                                         "y_fuel_consumption_per_capita_2010",      
                                         "y_wghtd_mean_commute_time" 
                            )
                            ,measure = list(c("cluster2","cluster3",
                                              "cluster4")),
                            ,variable.name = c("tipo_cluster")
                            ,value.name = c("cluster_num")) %>% 
  data.table::melt(data = .
                   ,id.vars = c('i_code_urban_concentration'
                                ,"tipo_cluster","cluster_num"
                   )
                   ,measure = list(c(cols_to_norm1,
                                     "y_fuel_consumption_per_capita_2010",      
                                     "y_wghtd_mean_commute_time")),
                   ,variable.name = c("urb_form_names")
                   ,value.name = c("urb_form_vars"))


df_plot %>% 
  .[tipo_cluster == "cluster2",] %>% 
  ggplot()+
  geom_boxplot(aes(x = as.character(cluster_num)
                   , y = urb_form_vars))+
  facet_wrap(~urb_form_names
             ,scales = "free",ncol = 3)
df_plot %>% 
  .[tipo_cluster == "cluster3",] %>% 
  ggplot()+
  geom_boxplot(aes(x = as.character(cluster_num)
                   , y = urb_form_vars))+
  facet_wrap(~urb_form_names
             ,scales = "free",ncol = 3)
df_plot %>% 
  .[tipo_cluster == "cluster4",] %>% 
  ggplot()+
  geom_boxplot(aes(x = as.character(cluster_num)
                   , y = urb_form_vars))+
  facet_wrap(~urb_form_names
             ,scales = "free",ncol = 3)

```
```{r ,echo=TRUE}

df_raw[cluster3==3]$i_name_urban_concentration
df_raw[cluster4==4]$i_name_urban_concentration

df_raw[cluster4==1]$i_name_urban_concentration
```

## density vs espraiamento
```{r ,echo=FALSE}
df_raw %>% 
  ggplot()+
  geom_jitter(aes(x = x_avg_cell_distance
                  , y = x_density_pop_05km_total_2014
                  , color = as.character(cluster2)
                  , size = x_pop_2015))+
  scale_x_log10() +
  scale_y_log10() + 
  labs(x = "avg_cell_distance: espraiamento"
       , y = "exp. density (5m)"
       , size = "population"
       , color = "Cluster")

```

```{r ,echo=FALSE}
df_raw %>% 
  ggplot()+
  geom_jitter(aes(x = x_avg_cell_distance
                  , y = x_density_pop_05km_total_2014
                  , color = as.character(cluster3)
                  , size = x_pop_2015))+
  scale_x_log10() +
  scale_y_log10() + 
  labs(x = "avg_cell_distance: espraiamento"
       , y = "exp. density (5km)"
       , size = "population"
       , color = "Cluster")

```

```{r ,echo=FALSE}
df_raw %>% 
  ggplot()+
  geom_jitter(aes(x = x_avg_cell_distance
                  , y = x_density_pop_05km_total_2014
                  , color = as.character(cluster4)
                  , size = x_pop_2015))+
  scale_x_log10() +
  scale_y_log10() + 
  labs(x = "avg_cell_distance: espraiamento"
       , y = "exp. density (05km)"
       , size = "population"
       , color = "Cluster")

```


# 6 saving

```{r ,echo=TRUE}
data.table::fwrite(x = df_raw
                   , file = "../../../../data/urbanformbr/pca_regression_df/cluster_output/cluster_1.csv")

```
